Isomorphisms as Groupoids: \\ Homotopy Type Theory with Fractional Types

### 2013-10-14 f.tex
f.tex with lots of interesting questions about pointed spaces

\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{url}
\usepackage{stmaryrd}
\usepackage{color, colortbl}
\usepackage{courier}
\usepackage{graphicx}

\newcommand{\alt}{~|~}
\newcommand{\leftv}[1]{\textsf{left}~#1}
\newcommand{\rightv}[1]{\textsf{right}~#1}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\identlp}{\mathit{identl}_+}
\newcommand{\identrp}{\mathit{identr}_+}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{identl}_*}
\newcommand{\identrt}{\mathit{identr}_*}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\distz}{\mathit{dist}_0}
\newcommand{\factorz}{\mathit{factor}_0}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\idc}{\mathit{id}}
\newcommand{\symc}[1]{\mathit{sym}~#1}
\newcommand{\proves}{\vdash}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2\\\end{array}}
{\begin{array}{l}#3\\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}

@@

We review our language $\Pi$ providing the necessary background and context
for our higher-order extension.\footnote{The presentation in this section
  focuses on the simplest version of $\Pi$. Other versions include recursive
  types and trace operators but these extensions are orthogonal to the
  higher-order extension emphasized in this document.} Even though the
semantics of $\Pi$ could be explained using plain sets, we express it using
the groupoid terminology to faciliate the transition to fractionals and
higher-order functions.

@@

\paragraph*{Data.}
We view $\Pi$ as having two levels: it has traditional values classified by
ordinary types:
\[\begin{array}{lrcl} 
\textit{values}, v &::=& () \alt \leftv{v} \alt \rightv{v} \alt (v,v) \\
\textit{value types}, b &::=& 0 \alt 1 \alt b+b \alt b * b
\end{array}\]

@@

\begin{table}[t]
\hrule
\smallskip
\begin{center} 
\Rule{}
{}
{\jdg{}{}{\idc : b \iso b}}
{}
\quad 
\Rule{}
{\jdg{}{}{c : b_1 \iso b_2}}
{\jdg{}{}{\symc{c} : b_2 \iso b_1}}
{}
\quad
\Rule{}
{\jdg{}{}{c_1 : b_1 \iso b_2} \quad c_2 : b_2 \iso b_3}
{\jdg{}{}{c_1 \fatsemi c_2 : b_1 \iso b_3}}
{}
\medskip 
\Rule{}
{\jdg{}{}{c_1 : b_1 \iso b_2} \quad c_2 : b_3 \iso b_4}
{\jdg{}{}{c_1 \oplus c_2 : b_1 + b_3 \iso b_2 + b_4}}
{}
\quad 
\Rule{}
{\jdg{}{}{c_1 : b_1 \iso b_2} \quad c_2 : b_3 \iso b_4}
{\jdg{}{}{c_1 \otimes c_2 : b_1 * b_3 \iso b_2 * b_4}}
{}
\end{center}
\caption{Combinators\label{pi-combinators}}
\smallskip
\hrule
\end{table}

@@

The language presented above, at the type level, models a \emph{semiring}
(occasionally called a \emph{rig}) where we replace equality by isomorphism.
Semantically, $\Pi$ models a \emph{bimonoidal category} whose simplest
example is the category of finite sets and bijections. In that
interpretation, each value type denotes a finite set of a size calculated by
viewing the types as natural numbers and each combinator $c : b_1 \iso b_2$
denotes a bijection between the sets denoted by $b_1$ and $b_2$.

%%%%%%%%%%%%%%%%%
\subsection{Operational Semantics}

Operationally, the semantics of $\Pi$ is given using two mutually recursive
interpreters: one going forward and one going backwards. The use of $\symc{}$
switches control from one evaluator to the other.  We state without proof
that the evaluation of well-typed combinators always terminates and that
$\Pi$ is logically reversible, i.e., that for all combinators $c : b_1 \iso
b_2$ and values $v_1 : b_1$ and $v_2 : b_2$ we have the forward evaluation of
$c v_1$ produces $v_2$ iff the backwards evaluation of $c v_2$ produces
$v_1$.

@@

In more detail, motivate and introduce pointed types here, add
an element $\bot_A$ to every type $A$ and express the groupoids using a
pointed type and an explicit collection of discrete paths (not indexed by
anything at this point).

>>>>

\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{url}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{color, colortbl}
\usepackage{courier}
\usepackage{graphicx}

\newcommand{\amr}[1]{\fbox{Amr says:} \textbf{#1}}

\newcommand{\union}{\cup}
\newtheorem{definition}{Definition}
\newcommand{\refl}{\mathit{refl}}
\renewcommand{\path}{\leadsto}
\renewcommand{\circ}{\circlearrowleft}
\newcommand{\alt}{~|~}
\newcommand{\dt}[1]{\llbracket #1 \rrbracket}
\newcommand{\leftv}[1]{\textsf{left}~#1}
\newcommand{\rightv}[1]{\textsf{right}~#1}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\identlp}{\mathit{identl}_+}
\newcommand{\identrp}{\mathit{identr}_+}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{identl}_*}
\newcommand{\identrt}{\mathit{identr}_*}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\distz}{\mathit{dist}_0}
\newcommand{\factorz}{\mathit{factor}_0}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\idc}{\mathit{id}}
\newcommand{\wedgesum}{\vee}
\newcommand{\smashproduct}{\wedge}
\newcommand{\symc}[1]{\mathit{sym}~#1}
\newcommand{\proves}{\vdash}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2\\\end{array}}
{\begin{array}{l}#3\\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}

@@

We review our language $\Pi$ providing the necessary background and context
for our higher-order extension.\footnote{The presentation in this section
  focuses on the simplest version of $\Pi$. Other versions include recursive
  types and trace operators but these extensions are orthogonal to the
  higher-order extension emphasized in this document.} Even though the
semantics of $\Pi$ could be explained using plain sets, we express it using
the groupoid terminology to facilitate the transition to fractionals and
higher-order functions.

@@

\paragraph*{Data.}
We view $\Pi$ as having two levels: it has traditional values classified by
ordinary types:
\[\begin{array}{rcl} 
\textit{values}, v &::=& () \alt \leftv{v} \alt \rightv{v} \alt (v,v) \\
\textit{value types}, b &::=& 0 \alt 1 \alt b+b \alt b * b
\end{array}\]

@@

\begin{table}[t]
\hrule\medskip
\begin{center} 
\Rule{}
{}
{\jdg{}{}{\idc : b \iso b}}
{}
\quad 
\Rule{}
{\jdg{}{}{c : b_1 \iso b_2}}
{\jdg{}{}{\symc{c} : b_2 \iso b_1}}
{}
\quad
\Rule{}
{\jdg{}{}{c_1 : b_1 \iso b_2} \quad c_2 : b_2 \iso b_3}
{\jdg{}{}{c_1 \fatsemi c_2 : b_1 \iso b_3}}
{}
\\ \bigskip
\Rule{}
{\jdg{}{}{c_1 : b_1 \iso b_2} \quad c_2 : b_3 \iso b_4}
{\jdg{}{}{c_1 \oplus c_2 : b_1 + b_3 \iso b_2 + b_4}}
{}
\quad 
\Rule{}
{\jdg{}{}{c_1 : b_1 \iso b_2} \quad c_2 : b_3 \iso b_4}
{\jdg{}{}{c_1 \otimes c_2 : b_1 * b_3 \iso b_2 * b_4}}
{}
\end{center}
\caption{Combinators\label{pi-combinators}}
\medskip\hrule
\end{table}

@@

@@

The language presented above, at the type level, models a \emph{commutative
  semiring} (occasionally called a \emph{commutative rig}) where we replace
equality by isomorphism.  Semantically, $\Pi$ models a \emph{bimonoidal
  category} whose simplest example is the category of finite sets and
bijections. In that interpretation, each value type denotes a finite set of a
size calculated by viewing the types as natural numbers and each combinator
$c : b_1 \iso b_2$ denotes a bijection between the sets denoted by $b_1$ and
$b_2$. Instead of presenting such a semantics, we present a slightly more
involved semantics using discrete groupoids. In that semantics, the
denotation of each type will consist of (i) a pointed set, and (ii) a trivial
equivalence relation relating each element to itself by a \emph{path}. The
pointed sets are constructed by adjoining, to each ``regular'' set, an
element $\circ$ which topologically corresponds to the
\emph{circle}. Significantly, using the Euler characteristic as a generalized
notion of cardinality, a set containing $n$ ``regular'' elements and a circle
$\circ$ has cardinality $n$ and not $n+1$.\footnote{Generally speaking, this
  generalized cardinality based on the Euler characteristic counts the number
  of ``connected components'' of a set. Consider a set with $n$ points and
  various paths connecting them. We start by counting all the points giving
  us cardinality $n$. Then for every path, we decrease the cardinality by one
  to avoid double-counting connected elements. Then we add one again for
  every path that forms a loop because the path forming the loop did not
  decrease the number of connected components. It is easy to see that a point
  with a path that connects it to itself would have generalized cardinality
  0. Moreover the generalized cardinality of the disjoint union of two sets
  is their sum and the generalized cardinality of the product of two sets is
  their product.}

\begin{definition}[Denotation of types]
Each type $b$ is mapped to a groupoid $(\dt{b}, \circ_b, \mathit{paths}_b)$
where $\dt{b}$ is a pointed set with distinguished point $\circ_b$ and
$\mathit{paths}_b$ is a collection of paths, $v_1 \path v_2$, between elements
$v_1$ and $v_2$ of $\dt{b}$. The map is defined as follows:
\[\begin{array}{rcl@{\qquad}l}
\dt{0} &=& (\{ \circ_0 \}, \circ_0, \{ \circ_0 \path \circ_0 \}) &
          (\mbox{circle}) \\
\dt{1} &=& (\{ \bullet, \circ_1 \}, \circ_1, 
           \{ \circ_1 \path \circ_1, \bullet \path \bullet \}) &
          (\mbox{0-sphere})\\
\dt{(b_1 + b_2)} &=& \dt{b_1} \wedgesum \dt{b_2} & 
          (\mbox{wedge~sum}) \\
\dt{(b_1 * b_2)} &=& \dt{b_1} \smashproduct \dt{b_2} & 
          (\mbox{smash~product})
\end{array}\]
The empty type is mapped to the topological circle which is a point that is
connected to itself. The unit type is mapped to the topological 0-sphere
which is a pair of points with the discrete topology. The operation
$\wedgesum$ is the \emph{wedge sum} which is a regular disjoint union with a
new distinguished point $\circ_{b_1+b_2}$ and additional paths consisting of
the equivalence closure of the identifications $\circ_{b_1} \path
\circ_{b_1+b_2}$ and $\circ_{b_2} \path \circ_{b_1+b_2}$. The operation
$\smashproduct$ is the \emph{smash product} which is a regular product with a
new distinguished point $\circ_{b_1*b_2}$ and additional paths consisting of
the equivalence closure of the identifications $(x,\circ_{b_2}) \path
\circ_{b_1*b_2}$ and $(\circ_{b_1},y) \path \circ_{b_1*b_2}$ for all $x \in
\dt{b_1}$ and $y \in \dt{b_2}$.
\end{definition}

We quickly check that the desired isomorphisms hold at the level of types,
e.g., the isomorphism $b * 0 \iso 0$ requires that the smash product of any
groupoid with the circle be indistinguishable from the circle.

The semantics of $\Pi$ is given using two mutually recursive interpreters:
one going forward and one going backwards. The use of $\symc{}$ switches
control from one evaluator to the other.

\[\begin{array}{rcl}
\mathit{eval} &:& (a \iso b) \rightarrow \dt{a} \rightarrow \dt{b} \\
\mathit{eval}~\idc~a &=& a \\
\mathit{eval}~(\symc{f})~b &=& \mathit{evalR}~f~b \\
\mathit{eval}~(f \fatsemi g)~a &=& \mathit{eval}~g~(\mathit{eval}~f~a) \\
\mathit{eval}~(f \otimes g)~(a,b) &=& (\mathit{eval}~f~a, \mathit{eval}~g~b) \\
\mathit{eval}~(f \oplus g)~(\leftv{a}) &=& \leftv{(\mathit{eval}~f~a)} \\
\mathit{eval}~(f \oplus g)~(\rightv{b}) &=& \rightv{(\mathit{eval}~g~b)} \\
\mathit{eval}~\identlp~(\leftv{\circ}) &=& \circ \\
\mathit{eval}~\identlp~(\rightv{a}) &=& a \\
\mathit{eval}~\identrp~a &=& \rightv{a} \\
\mathit{eval}~\swapp~(\leftv{a}) &=& \rightv{a} \\
\mathit{eval}~\swapp~(\rightv{b}) &=& \leftv{b} \\
\mathit{eval}~\assoclp~(\leftv{a}) &=& \leftv{(\leftv{a})} \\
\mathit{eval}~\assoclp~(\rightv{(\leftv{b})}) &=& \leftv{(\rightv{b})} \\
\mathit{eval}~\assoclp~(\rightv{(\rightv{c})}) &=& \rightv{c} \\
\mathit{eval}~\assocrp~(\leftv{(\leftv{a})}) &=& \leftv{a} \\
\mathit{eval}~\assocrp~(\leftv{(\rightv{b})}) &=& \rightv{(\leftv{b})} \\
\mathit{eval}~\assocrp~(\rightv{c}) &=& \rightv{(\rightv{c})} \\
\mathit{eval}~\identlt~((), a) &=& a \\
\mathit{eval}~\identrt~a &=& ((), a) \\
\mathit{eval}~\swapt~(a,b) &=& (b,a) \\
\mathit{eval}~\assoclt~(a,(b,c)) &=& ((a,b),c) \\
\mathit{eval}~\assocrt~((a,b),c)  &=& (a,(b,c)) \\
\mathit{eval}~\distz~(\rightv{c}, a) &=& \rightv{(c, a)} \\
\mathit{eval}~\factorz~(\leftv{(b, a)}) &=& (\leftv{b}, a) \\
\mathit{eval}~\dist~(\leftv{b}, a) &=& \leftv{(b, a)} \\
\mathit{eval}~\dist~(\rightv{c}, a) &=& \rightv{(c, a)} \\
\mathit{eval}~\factor~(\leftv{(b, a)}) &=& (\leftv{b}, a) \\
\mathit{eval}~\factor~(\rightv{(c, a)}) &=& (\rightv{c}, a) \\
\end{array}\]
\[\begin{array}{rcl}
\mathit{evalR} &:& (a \iso ) \rightarrow b \rightarrow a \\
\mathit{evalR}~\idc~a &=& a \\
\mathit{evalR}~(\symc{f}) b &=& \mathit{eval}~f b \\
\mathit{evalR}~(f \fatsemi g) a &=& \mathit{evalR}~f (\mathit{evalR}~g a) \\
\mathit{evalR}~(f \otimes g) (a,b) &=& 
  (\mathit{evalR}~f a, \mathit{evalR}~g b) \\
\mathit{evalR}~(f \oplus g) (\leftv{a}) &=& \leftv{(\mathit{evalR}~f a)} \\
\mathit{evalR}~(f \oplus g) (\rightv{b}) &=& \rightv{(\mathit{evalR}~g b)} \\
\mathit{evalR}~\identlp~a &=& \rightv{a} \\
\mathit{evalR}~\identrp~(\rightv{a}) &=& a \\
\mathit{evalR}~\swapp~(\leftv{a}) &=& \rightv{a} \\
\mathit{evalR}~\swapp~(\rightv{b}) &=& \leftv{b} \\ 
\mathit{evalR}~\assoclp~(\leftv{(\leftv{a})}) &=& \leftv{a} \\
\mathit{evalR}~\assoclp~(\leftv{(\rightv{b})}) &=& \rightv{(\leftv{b})} \\
\mathit{evalR}~\assoclp~(\rightv{c}) &=& \rightv{(\rightv{c})} \\
\mathit{evalR}~\assocrp~(\leftv{a}) &=& \leftv{(\leftv{a})} \\
\mathit{evalR}~\assocrp~(\rightv{(\leftv{b})}) &=& \leftv{(\rightv{b})} \\
\mathit{evalR}~\assocrp~(\rightv{(\rightv{c})}) &=& \rightv{c} \\
\mathit{evalR}~\identlt~a &=& ((), a) \\
\mathit{evalR}~\identrt~((), a) &=& a \\
\mathit{evalR}~\swapt~(a,b) &=& (b,a) \\
\mathit{evalR}~\assoclt~((a,b),c)  &=& (a,(b,c)) \\
\mathit{evalR}~\assocrt~(a,(b,c)) &=& ((a,b),c) \\
\mathit{eval}~\distz~(\rightv{c}, a) &=& \rightv{(c, a)} \\
\mathit{eval}~\factorz~(\leftv{(b, a)}) &=& (\leftv{b}, a) \\
\mathit{evalR}~\dist~(\leftv{(b, a)}) &=& (\leftv{b}, a) \\
\mathit{evalR}~\dist~(\rightv{(c, a)}) &=& (\rightv{c}, a) \\
\mathit{evalR}~\factor~(\leftv{b}, a) &=& \leftv{(b, a)} \\
\mathit{evalR}~\factor~(\rightv{c}, a) &=& \rightv{(c, a)}
\end{array}\]

We state without proof that the evaluation of well-typed combinators always
terminates and that $\Pi$ is logically reversible, i.e., that for all
combinators $c : b_1 \iso b_2$ and values $v_1 : b_1$ and $v_2 : b_2$ we have
the forward evaluation of $c v_1$ produces $v_2$ iff the backwards evaluation
of $c v_2$ produces $v_1$.
